Calculating life expectancy and exploring missingness

### 2016-11-08 Code/1_Explore_Data_v2.Rmd
adding in LE for 1 and 9 but not yet finished

dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")

dat2 <- dat2 %>% arrange(State2, Year3, Race2, Sex2, Age3)
```

@@

```{r calculate_life_expectancy}
list.life.tables <- vector(mode = "list", length = length(non.zero.stratums))
list.index <- 1

system.time(
for(stratum.id in non.zero.stratums) {

  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id), 
                                               num.ages.in.group = "nx", 
                                               death.counts = "total_deaths5", 
                                               population.counts = "Population")
  list.index <- list.index + 1
  }
)

@@

```{r add_le_to_dataset}
le <- rep(NA, length(non.zero.stratums))

for (i in 1:length(non.zero.stratums)) {
  le[i] <- list.life.tables[[i]]$e_x[1]

les <- data.frame(life.expectancy.birth = le, StateYearRaceSex = non.zero.stratums)

@@

Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth) %>%
                  rename(Pop_black = Pop_across_age,  le.birth.black = life.expectancy.birth)

White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth) %>%
                  rename(Pop_white = Pop_across_age,  le.birth.white = life.expectancy.birth)

@@

BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black)
```

@@

list.age.decomp.tables <- vector(mode = "list", length = dim(paired.ids)[1])

@@

  if(length(white.index) != 0 & length(black.index) != 0) {
    list.age.decomp.tables[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")
  } else {
    list.age.decomp.tables[[i]] <- "Couldn't compare life tables"
  }
}
```

@@

#takes about 20 mins
list.cod.decomp.tables <- vector(mode = "list", length = dim(paired.ids)[1])

@@

  if(length(white.index) != 0 & length(black.index) != 0) {
    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
                ungroup() %>%
                select(Age2, COD2, cod_prop_death5) %>%
                rename(prop.white = cod_prop_death5)

    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
                ungroup() %>%
                select(Age2, COD2, cod_prop_death5) %>%
                rename(prop.black = cod_prop_death5)
    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
    ct.both <- ct.both %>% arrange(COD2, Age2)

    list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                         life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                         decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                         cod.table = ct.both, 
                                                         age.colname.cod.table = "Age2", 
                                                         COD.colname.cod.table = "COD2", 
                                                         prop1.colname.cod.table = "prop.black", 
                                                         prop2.colname.cod.table = "prop.white")
  } else {
    list.cod.decomp.tables[[i]] <- "Couldn't compare life tables"
  }  
  #print(i)
}
```

>>>>

dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")
rm(dat3)
dat2 <- dat2 %>% arrange(State2, Year3, Race2, Sex2, Age3)
```

@@

```{r calculate_life_expectancy}
list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
#list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
list.index <- 1

system.time(
for(stratum.id in non.zero.stratums) {
#for(stratum.id in head(non.zero.stratums)) {

  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2),   
  #list.life.tables1[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
                                               num.ages.in.group = "nx", 
                                               death.counts = "total_deaths1", 
                                               population.counts = "Population")

  list.life.tables5[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
                                               num.ages.in.group = "nx", 
                                               death.counts = "total_deaths5", 
                                               population.counts = "Population")

  list.life.tables9[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
                                               num.ages.in.group = "nx", 
                                               death.counts = "total_deaths9", 
                                               population.counts = "Population")
  list.index <- list.index + 1
  print(list.index)
  }
)
```

@@

```{r add_le_to_dataset}
le <- rep(NA, length(non.zero.stratums))
le5 <- rep(NA, length(non.zero.stratums))
le9 <- rep(NA, length(non.zero.stratums))

for (i in 1:length(non.zero.stratums)) {
  le[i] <- list.life.tables[[i]]$e_x[1]
  le5[i] <- list.life.tables5[[i]]$e_x[1]
  le9[i] <- list.life.tables9[[i]]$e_x[1]
}

les <- data.frame(life.expectancy.birth = le, life.expectancy.birth.5 = le5, life.expectancy.birth.9 = le9, StateYearRaceSex = non.zero.stratums)

@@

Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth, life.expectancy.birth5, life.expectancy.birth9) %>%
                  rename(Pop_black = Pop_across_age,  
                         le.birth.black = life.expectancy.birth, 
                         le.birth.black5 = life.expectancy.birth5, 
                         le.birth.black9 = life.expectancy.birth9)

White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, life.expectancy.birth, life.expectancy.birth5, life.expectancy.birth9) %>%
                  rename(Pop_white = Pop_across_age,  
                         le.birth.white = life.expectancy.birth,
                         le.birth.white5 = life.expectancy.birth5,
                         le.birth.white9 = life.expectancy.birth9)

@@

BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black,
                                    WBgap5 = le.birth.white5 - le.birth.black5,
                                    WBgap9 = le.birth.white9 - le.birth.black9)
```

@@

list.age.decomp.tables <- list.age.decomp.tables5 <- list.age.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])

@@

  if(length(white.index) != 0 & length(black.index) != 0) {
    list.age.decomp.tables[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")

    list.age.decomp.tables5[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")

    list.age.decomp.tables9[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")
  } else {
    list.age.decomp.tables[[i]] <- list.age.decomp.tables5[[i]] <- list.age.decomp.tables9[[i]] <-"Couldn't compare life tables"
  }
}
```

@@

#takes about 20 mins
list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])

@@

  if(length(white.index) != 0 & length(black.index) != 0) {
    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
                ungroup() %>%
                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
                rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)

    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
                ungroup() %>%
                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
                rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)
    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
    ct.both <- ct.both %>% arrange(COD2, Age2)

    list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                         life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                         decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                         cod.table = ct.both %>% select(Age2, COD2, prop.white5, prop.black5), 
                                                         age.colname.cod.table = "Age2", 
                                                         COD.colname.cod.table = "COD2", 
                                                         prop1.colname.cod.table = "prop.black", 
                                                         prop2.colname.cod.table = "prop.white")
  } else {
    list.cod.decomp.tables[[i]] <- "Couldn't compare life tables"
  }  
  #print(i)
}
```
